Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)

### 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process HBCC DLPFC RNASeq data

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
                                 parentId ="syn9846623",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9846623';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitFixedEffectModel <- function(varsToFit, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = plyr::ldply(1:dim(lcpm)[1],
                         .fun = function(i, lcpm, designMat, voomWeights){
                           # Fit a linear models for every gene
                           data = cbind(data.frame(Gx = lcpm[i,]),
                                        designMat, 
                                        weight = voomWeights[i,])

                           formula = paste(setdiff(colnames(data), c('Gx', 'weight')), collapse = '+') %>%
                             paste('Gx ~',., collapse = ' ')

                           mdl = stats::lm(formula = formula, 
                                           data = data, 
                                           weights = weight)

                           data.frame(ensembl_gene_id = rownames(lcpm)[i], BIC = BIC(mdl))
                         }, 
                         lcpm, designMat, voomWeights, 
                         .parallel = F,
                         .paropts = list(.packages = c('dplyr')))
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = plyr::llply(1:length(covarsToSelect),
                              .fun = function(i, covarsToSelect, adjustCovars, lcpm, cnt, covars){

                                # Fit a mixed effect linear model and get all BIC
                                modelBIC = fitFixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                               lcpm, cnt, covars) %>%
                                  plyr::rename(c('BIC' = covarsToSelect[i]))

                                return(modelBIC)
                              },
                              covarsToSelect, adjustCovars, lcpm, cnt, covars, 
                              .parallel = F,
                              .paropts = list(.export = c('fitFixedEffectModel')))

    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx.Tissue')
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'RibozeroBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.", "Mean_CV", "Fragment_Length_StdDev",
                  "Mapped_Pairs", "End_1_Mismatch_Rate", "End_2_Mismatch_Rate", "Mapped", "IntragenicRate",
                  "ExonicRate", "IntergenicRate", "IntronicRate", "rRNARate", "GenesDetected",
                  "TranscriptsDetected", "ExpProfEfficiency", "PercentMapped")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=8}
# Covariates to adjust
postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')

# Post adjusted design matrix
DESIGN = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = limma::voom(NEW.COUNTS, design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

  normalizationType	= 'CQNandMS'
)

# Code
CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store filtered counts
NEW.COUNTS %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

@@

### Custom Normalisation
Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue).

Based on the model selection analysis here (syn9907541), following variables were controlled for differential expression: 
    Gender, RIN, IntronicRate
```{r custom.normalisation, results='asis'}
adjust.covars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', ')))

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=8}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store filtered counts
NEW.COUNTS %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_nCounts.tsv', sep = '\t', row.names=F, quote=F)
NCOUNT_OBJ = File('CMC_ACC_DLPFC_nCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
annotations(NCOUNT_OBJ) = annotations(CODE)
annotations(NCOUNT_OBJ)$dataSubType = 'estimatedCounts'
NCOUNT_OBJ = synStore(NCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (estimated)                         |  `r NCOUNT_OBJ$properties$id`    |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
